Statically Typed String Sanitation Inside a Python (Technical Report)

### 2014-12-13 sanitation-psp14/tr/strings_chapter.tex
Saturday/sunday.

\newtheorem{tr}{TR Item}
\renewcommand{\thetr}{\Alph{tr}} % "letter-numbered" theorems
\newtheorem{trexam}[tr]{TR Example}
\newtheorem{trthm}[tr]{TR Theorem}
\newtheorem{trlem}[tr]{TR Lemma}
\newtheorem{trdef}[tr]{TR Definition}
\newtheorem{trconj}[tr]{TR Conjecture}

@@

\begin{trlem}[Leading characters are in the head] \label{trlemmax}
If $c_1 \cdot c_2 \cdot ... \cdot c_m \in \lang{r}$, then
$c_1 \in \lhead{r}$.
\end{trlem}
\begin{proof}
By structural induction on $r$...
\end{proof}

\begin{trthm}[Correctness of \textsf{ltail}]
If $s \in \lang{r}$ then $s \in \lang{\lhead{r}} \cdot \lang{\ltail{r}}$.
\end{trthm}
\begin{proof}
The proof proceeds by structural induction on $r$.
In each case, we demonstrate three facts:
\begin{itemize}
  \item First, that the string $s = a \cdot b$ is a concatenation of two substrings (each of which may be $\epsilon$).
  \item Second, $a \in \lang{\lhead{r}}$ (or, equivalently, $s \in \lang{\lhead{r}}$ letting $b = \epsilon$).
  \item Third, there exists a $c \in \lhead{r}$ such that $a = c \cdot a'$ and $a' \cdot b \in \lang{\delta_c(r)}$, or else $s = a$. 
\end{itemize}

Throughout, we consider $\epsilon$ as idenetity under all operators for expressions and $\cdot$ for strings.

@@

\item \textbf{Case $r = c$ for $c \in Sigma$}.
Note that if $s \in \lang{c}$ then $s = c$.
Note that $\lhead{c} = \lhead{c, \epsilon} = c$, and $c \in \lang{c}$.
So $c \in \lang{c}$, which is sufficient since $\epsilon \in \ltail{r}$.

\item \textbf{Case $r = r_1 \cdot r_2$}.
Suppose $s \in \lang{r}$.
We may decompose $s$ as $s_1 \cdot ... \cdot s_n$ such that $s_1 = c_1 \cdot ... \cdot c_m \in r_1$.
Note that $c_1 \in \lhead{r_1}$ by \ref{trlemmax}.
Let $a' = c_2 \cdot ... \cdot c_m$.
It suffices to show that $a' \cdot b \in \lang{\delta_c(r)}$.
Since $s = c_1 \cdot a' \cdot b$ and $s \in \lang{r}$, it follows by the definition of derivative that $a' \cdot b \in \delta_{c_1}(r)$.

\item \textbf{Case $r = r_1 + r_2$}.
Suppose $s \in \lang{r_1 + r_2}$
so that $s \in \lang{r_1}$ or $s \in \lang{r_2}$.
In either case, the result follows by induction.

\item \textbf{Case $r = q^*$}.
Note that $s = \epsilon$ or else $s = s_1 \cdot s_2 \cdot ... s_n$ where each $s_i \in \lang{q}$\footnote{intuitively, $s$ is either empty or a finitary concatenation of strings matching $q$.}.

Suppose $s = \epsilon$.
Note that $\lhead{r} = \lhead{q^*} = \lhead{q^*, \epsilon} = \lhead{\epsilon, \epsilon} + h' = \epsilon + h'$.
Therefore, $s \in \lhead{r}$.

In the other case, suppose $s = s_1 \cdot ... \cdot s_n$ where each $s_i \in \lang{q}$.
Let $a = s_1$ and $b = s_2 \cdot ... s_n$, so that $s = a \cdot b$.
Note that $a = c_1 \cdot ... \cdot c_n$.
We will show that $c_1 \in \lang{\lhead{q^*}}$ and that $(c_2 \cdot ... c_n) \cdot b \in \delta_{c_1}(q^*)$.

The latter property is trivial. Note that $s = c_1 \cdot c_2 \cdot ... \cdot c_n \cdot b \in \lang{q^*}$.
The result follows immediately from the definition of derivative.

What remains to be shown is that $c_1 \in \lang{q^*}$.
Note that $\lhead{q^*} = \epsilon + \lhead{q, \epsilon}$.
Recall that $a = s_1$ where $s_1 \in \lang{q}$.
Also recall that $a= c_1 \cdot ... \cdot c_n$.
Therefore, $c_1 \in \lhead{q}$ by \ref{trlemmax}.

\begin{trexam}[All the heads of all the tails can be more than one head and tail]\label{exam:headsandtails}
$r \not = \lhead{r} \cdot \ltail{r}$.
\end{trexam}
\begin{proof}
A simple counter-example is $ab + cd$. 
Note that $\lhead{ab + cd} = a+c$ and $\ltail{ab + cd} = b + d$. 
Therefore, $\{ad,bc\} \subset \lang{\lhead{ab + cd} \cdot \ltail{ab + cd}}$ even though neither of these is in $\lang{r}$.
\end{proof}

Example~\ref{exam:headsandtails} does not imply a counter-example to type soundness because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
property required for soundness. Still, in a production implementation, it will make sense to massage the definitions of $\lhead{r}$ and $\ltail{r}$ so that type information is not unnecessarily lost during substring operations.

@@

@@

\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
~
\begin{itemize}

\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
\item For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
\item Regular languages are closed under reversal.
\end{itemize}
\end{lem}

@@

\begin{lem}
  If $\emptyset \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
\end{lem}
\begin{proof}
  The proof proceeds by induction on the derivation of the premise.
  The only non-trivial cases (those which require more than an appeal to inversion) are S-T-Case, S-T-Replace and S-T-Concat.

In the S-T-Case case, note that ${\sf lhead}$ and ${\sf ltail}$ are total functions for well-formed regular expressions to well-formed regular expressions.

In the S-T-Concat case, note that Lemma~\ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.

In the S-T-Replace case, it suffices to show that $\lreplace{r}{r_1}{r_2}$ is a regular expression assuming (inductively) that $r, r_1$ and $r_2$ are all regular expressions.
This follows from the Closure proposition.
\end{proof}

@@

\begin{trconj}
if $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ such that $v \ \val$.
\end{trconj}
%\begin{trthm}[Termination of small step semantics]
%\end{trthm}
%\begin{proof}
We do not develop the full proof here, but note that the simply typed lambda calculus terminates.
For the string fragment, observe that the S-T- rules do not add any non-trivial binding structure because substitutions
$[e/x]e'$ may only occur in the special case where $e = \rstr{s}$, so that the length of the term
never increases and the number of free variables strictly decreases. Therefore, the standard normalization
argument proceeds without complication after fixing an evaluation order for the compatibility rules 
(all our other proofs are agnostic to evaluation order).
%\end{proof}

\begin{trlem}[Canonical Forms]\label{cfgen}
Suppose $v \ \val$.%or $e \sreduces v$.

If $\emptyset\vdash v : \stringin{r}$ then $v = \rstr{s}$.

If $\emptyset\vdash v : \sigma \rightarrow \sigma'$ then $v = \lambda x . e'$ for some $e'$.
\end{trlem}
\begin{proof}
By inspection of valuation and typing rules.
\end{proof}

@@

\begin{lem}[Canonical Forms for the String Fragment of $\lambdas$]\label{thm:cfs}
  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
\end{lem}
\begin{proof}
This fact follows directly from Lemma~\ref{cfgen}.
\end{proof}

\begin{trlem}[Progress of small step semantics.]
If $\emptyset\vdash e : \sigma$ either $e \ \val$ or $e \mapsto e'$ for some $e'$.
\end{trlem}
\begin{proof}
The proof proceeds by induction on the derivation of $\emptyset\vdash e : \sigma$.
\begin{itemize}[label=$ $,itemsep=1ex]
\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of type safety for
the simply typed lambda calculus.

\item \textbf{S-T-Stringin-I}. Suppose $\emptyset\vdash \rstr{s} : \stringin{s}$.
The $\rstr{s} \ \val$ by SS-E-RStr.

\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{s}$.
By inversion and induction, $e_1 \mapsto e_1'$ or $e_1 \ \val$ and similarly
for $e_2$. 
If $e_1$ steps, then SS-E-Concat-Left applies and so
$\rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$. Similarly, if $e_2$ steps
then $e$ steps by SS-E-Concat-Right.

In the remaining case, $e_1 \ \val$ and $e_2 \ \val$.
But then it follows by Canonical Forms that $e_1 = \rstr{s_1}$
and $e_2 = \rstr{s_2}$.
Finally, by SS-E-Concat, \\ $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.

\def\mline#1{\par\hspace*{-\leftmargin}\parbox{\textwidth}{\[#1\]}}
\item \textbf{S-T-Case}.
Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$. 
By inversion,\\ $\emptyset\vdash e_1 : \stringin{r}$.
%\begin{align}
%&\vdash e_1 : \stringin{r} \\
%&\vdash e_2 : \sigma \\
%x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma
%\end{align}
%Actually this isn't even necessary...
%By (2), $e_2 \mapsto e_2'$ or $e_2 \val$. 
%In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
From this fact, induction, and Canonical Forms it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
In the former case, $e$ steps by S-E-Case-Left.
In the latter case, note that $s = \epsilon$ or $s = at$ for some string $t$.
If $s = \epsilon$ then $e$ steps by S-E-Case-$\epsilon$-Val,
and if $s = at$ the $e$ steps by S-E-Case-Concat.

\item \textbf{S-T-Replace}.
Suppose $e = \rreplace{r}{e_1}{e_2}$ and $\emptyset\vdash e : \stringin{r'}$ where, by inversion of S-T-Replace,
\begin{align}
&\emptyset\vdash e_1 : \stringin{r_1} \\
&\emptyset\vdash e_2 : \stringin{r_2} \\
&\lreplace{r}{r_1}{r_2} = r'
\end{align}

By (1), inversion and induction $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$,
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-Replace-Left.
Similarly, if $e_2$ steps then $e$ steps by SS-E-Replace-Right.
The only remaining case is where $e_1 \ \val$ and also $e_2 \ \val$.
But then by Canonical Forms, $e_1 = \strin{s_1}$ and $e_2 = \strin{s_2}$.
Therefore, $e \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ by SS-E-Replace.

\item \textbf{S-T-SafeCoerce}.
Suppose that $\emptyset\vdash \rcoerce{r}{e_1} : \stringin{r}$.
By inversion of S-T-SafeCoerce,
%\begin{align}
$\emptyset\vdash e_1 : \stringin{r'}$ for $\lang{r'} \subseteq \lang{r}$.% \\
%&\lang{r'} \subseteq \lang{r}
%\end{align}
By induction, $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$.
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-SafeCoerce-Step.
Otherwise, $e_1 \ \val$ and by Canonical Forms $e_1 = \rstr{s}$.
In this case, $e = \rcoerce{r}{\rstr{s}} \mapsto \rstr{s}$ by SS-E-SafeCoerce.

\item \textbf{S-T-SafeCheck}
Suppose that $\emptyset\vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \stringin{r}$.
By inversion of S-T-Check:
\begin{align}
&\vdash e_0 : \stringin{r_0}  \\
x:\stringin{r} &\vdash e_1 : \sigma \\
&\vdash e_2 : \sigma \label{align6}
\end{align}

By (\ref{align6}) and induction,  $e_0 \mapsto e_0'$ or $e_0 \ \val$.
In the former case $e$ steps by SS-E-Check-StepRight.
Otherwise, $e_0 = \rstr{s}$ by Canonical Forms.
By Lemma~\ref{thm:regexprops}, either $s \in \lang{r_0}$ or $s \not\in\lang{r_0}$.
In the former case $e$ takes a step by SS-E-Check-Ok.
In the latter case $e$ takes a step by SS-E-Check-NotOk.

@@

\begin{trlem}[Preservation for Small Step Semantics]
If $\emptyset \vdash e : \sigma$ and $e \mapsto e'$ then $\emptyset \vdash e : \sigma$.
\end{trlem}
\begin{proof}

By induction on the derivation of $e \mapsto e'$.

@@

\item \textbf{SS-E-Concat-Left}.
Suppose $e = \rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$ and $e_1 \mapsto e_1'$.
By inversion of S-T-Concat, $\emptyset\vdash e_1 : \stringin{r_1}$ where \\$\emptyset\vdash e : \stringin{r_1r_2}$.
By induction, if $e_1 \mapsto e_1'$ then $\emptyset\vdash e_1' : \stringin{r_1}$.
Therefore, $\emptyset\vdash \rconcat{e_1'}{e_2} : \stringin{r_1r_2}$.

\item {\textbf{SS-E-Concat-Right}}. \todo{Similar to SS-E-Concat-Left.}

\item \textbf{SS-E-Concat}. Suppose $\emptyset\vdash\rconcat{\rstr{s_1}}{\rstr{s_2}} : \stringin{r_1r_2}$ and
$\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
Then by inversion $\emptyset\vdash\rstr{s_1} : \stringin{r_1}$ and similarly for $\rstr{s_2}$.
Therefore, $s_1 \in \lang{r_1}$ and\\ $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
that $s_1s_2 \in \lang{r_1r_2}$. Therefore, $\emptyset\vdash \rstr{s_1s_2} : \stringin{r_1r_2}$ by S-T-Rstr.

\item \textbf{S-E-Case-Left}.
Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3}$ and also that
$e \mapsto \strcase{e_1'}{e_2}{x,y.e_3}$
and $\emptyset \vdash e : \stringin{r}$.
By inversion of S-T-Case:
\begin{align}
\emptyset&\vdash e_1 : \stringin{r} \label{align7} \\
\emptyset&\vdash e_2 : \sigma \label{align8} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align9}
\end{align}
By (\ref{align7}) and the assumption that $e_1 \mapsto e_1'$, it follows by induction that $\emptyset\vdash e_1' : \stringin{r}$.
This fact together with (\ref{align8}) and (\ref{align9}) implies by S-T-Case that
$\emptyset\vdash\strcase{e_1'}{e_2}{x,y.e_3} : \sigma$.

>>>>

\newtheorem{tr}{TR Item}
\renewcommand{\thetr}{\Alph{tr}} % "letter-numbered" theorems
\newtheorem{trexam}[tr]{Example}
\newtheorem{trthm}[tr]{Theorem}
\newtheorem{trlem}[tr]{Lemma}
\newtheorem{trdef}[tr]{Definition}
\newtheorem{trconj}[tr]{Conjecture}
\newtheorem{ass}[tr]{Assumption}

@@

Throughout the following proofs, we abbreviate $s \in \lang{r}$ by $s \in r$.

\begin{trlem}[Leading characters are in the head] \label{trlemmax}
If $s = c_1  c_2 \cdots c_m \in \lang{r}$, then
$c_1 \in \lhead{r}$.
\end{trlem}
\begin{proof}
By structural induction on $r$.

If $r = x$ for some character $x$ then $s \in \lang{r}$ implies $s = x$.
By definition of lhead, $\lhead{s} = \lhead{s, \epsilon} = \lhead{x, \epsilon} = x$.
Finally, $x \in \lang{x}$.

If $r = r_1 \cdot r_2$ then $s \in \lang{r}$ implies $s = c_1 s_1 s_2$
such that $c_1 s_1 \in \lang{r_1}$. 
By definition of lhead, $\lhead{r_1 \cdot r_2} = \lhead{r_1, r_2} =
\cdots + \lhead{r_1, \epsilon} = \cdots + \lhead{r_1}$. Therefore, $r_1 \subseteq \lhead{r_1, r_2}$
and so it suffices to show that $s = c_1 s_1 s_2$ such that $c_1 \in \lhead{r_1}$,
\todo{this is why we break out the $c_1$ from the $s_1$, both here and going forward}
which holds by induction because $c_1 s_1 \in \lang{r_1}$.

If $r = r_1 + r_2$ then $s \in r$ implies either $s \in r_1$ or $s \in r_2$.
Without loss of generality suppose $s \in r_1$.
Then by induction $s = c_1 \cdots c_m$ such that $c_1 \in \lhead{r_1}$.
By definition, $\lhead{r} = \lhead{r_1, \epsilon} + \lhead{r_2, \epsilon} = \lhead{r_1} + \cdots$.
Therefore, $c_1 \in \lhead{r}$.

If $r = r_1^*$ then $s \in r$ implies that $s = s_1 \cdots s_k$ where $k \ge 0$ and $s_i \in \lang{r}$.
By definition of lhead, $\lhead{r_1^*} = \epsilon + \lhead{r_1}$.
If $s = \epsilon$, then $s \in \lang{\epsilon} \subseteq \lang{\lhead{r_1}}$.
If $s = s_1 \cdots s_k$, then $s_1 \in \lang{r}$ and so by induction $s_1 = c_1 \cdots c_k$ such that $c_1 \in \lang{\lhead{r}}$.
\end{proof}

\begin{trthm}[Correctness of \textsf{ltail}]
If $s \in \lang{r}$ then $s \in \lang{\lhead{r} \cdot \ltail{r}}$.
\end{trthm}
\begin{proof}
The proof proceeds by structural induction on $r$.
In each case, we identify sublanguages of the head and tail of $r$ such that 
$s$ is in the concatenation of the the sublanguages.
Throughout the proof, we use the fact that $\lang{r'} \subseteq \lang{r}$ and $\lang{q'} \subseteq \lang{q}$
implies $\lang{r' \cdot q'} \subseteq \lang{r \cdot q}$.

@@

\item \textbf{Case $r = c$ for $c \in Sigma$}.
By definition, if $s \in \lang{c}$ then $s = c$.
By definition of lhead and ltail, $\lhead{c} = \lhead{c, \epsilon} = c$
and $\lang{\epsilon} \subseteq \lang{\ltail{r}}$.
Therefore, $c \in \lang{c \cdot \epsilon} \subseteq \lang{\lhead{r} + \ltail{r}}$.

\item \textbf{Case $r = r_1 \cdot r_2$}.
Suppose $s \in \lang{r}$.
We may decompose $s$ as $s_1  \cdots s_n$ such that $s_1 = c_1 s_1' \in \lang{r_1}$.
Note that $c_1 \in \lhead{r_1}$ by \ref{trlemmax}.
Therefore, $c_1 \in \lhead{r_1 \cdot r_2}$ by definition of lhead.
It suffices to show that $s_1' s_2 \in \lang{\delta_{c_1}(r)}$.
Since $s = c_1 s_1' s_2\in \lang{r}$, it follows by the definition of derivative that $s_1' s_2 \in \delta_{c_1}(r)$.
Therefore, $s = c_1 s_1' s_2 \in \lang{ c_1 \cdot \delta_{c_1} } \subseteq \lang{ \lhead{r} \cdot \ltail{r}}$.

\item \textbf{Case $r = r_1 + r_2$}.
Suppose $s \in \lang{r_1 + r_2}$
so that $s \in \lang{r_1}$ or $s \in \lang{r_2}$.
Consider without loss of generality the case where $s \in \lang{r_1}$.
By induction, $s \in \lang{\lhead{r_1} \cdot \ltail{r_1}}$
Note that $\lhead{r_1} \subseteq \lhead{r}$
and $\ltail{r_1} \subseteq \ltail{r}$.
Therefore, $\lhead{r_1} \cdot \ltail{r_2} \subseteq \lhead{r_1} \cdot \ltail{r_1}$.

\item \textbf{Case $r = q^*$}.
Either $s$ is the empty string or else the $k^{th}$ unwinding (i.e., $s = s_1 \cdots s_k$ where $k > 0$ and $s_i \in \lang{q}$).

Suppose $s = \epsilon$.
By definition of lhead, $\lhead{r} = \lhead{q^*} = \lhead{q^*, \epsilon} = \lhead{\epsilon, \epsilon} + \cdots = \epsilon + \cdots$.
By definition of ltail, $\ltail{r} = \epsilon \cdots$.
From these facts we conclude that $s = \epsilon \in \lang{\epsilon \cdot \epsilon} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$.

Suppose instead that $s = s_1 \cdots s_k$ for $k < 0$ where $s_i \in \lang{q}$.
If $q = \epsilon$ then $s = \epsilon$ and the conclusion follows for the same reason as the epsilon subcase of the character case.
Otherwise, $s$ is not the empty string and so $s = c_1 s_1 s_2$ where $s_1$ and $s_2$ might be empty.
By the lemma above, $c_1 \in \lhead{s}$.
By the definition of derivative, $s_1s_2 \in \delta_{c_1}(s)$.
Therefore, $s \in \lang{\lhead{s} \cdot \delta_{c_1}(s)} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$, where the set inclusion follows by the definition
of lhead and ltail.
\end{itemize}
\end{proof}

\begin{trexam}[Decomposition is not Reversible]\label{exam:headsandtails}
$r \not = \lhead{r} \cdot \ltail{r}$.
\end{trexam}
\begin{proof}
A simple counter-example is $ab + cd$. 
Note that $\lhead{a\cdot b + c\cdot d} = a+c$ and $\ltail{a\cdot b + c\cdot d} = b + d$. 
Therefore, $\{ad,bc\} \subset \lang{\lhead{a\cdot b + cd\cdot } \cdot \ltail{a\cdot b + c\cdot d}}$ even though neither of these is in $\lang{r}$.
\end{proof}

Example~\ref{exam:headsandtails} does not imply a counter-example to type soundness because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
property required for soundness.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Substitution.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{ass}[Properties of Regular Languages.] \label{thm:regexprops}
~
\begin{enumerate}

\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
\item For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
\item Regular languages are closed under reversal.
\end{enumerate}
\end{ass}

@@

%\begin{lem}
%  If $\emptyset \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
%\end{lem}
%\begin{proof}
%  The proof proceeds by induction on the derivation of the premise.
%  The only non-trivial cases (those which require more than an appeal to inversion) are S-T-Case, S-T-Replace and S-T-Concat. 
%
%In the S-T-Case case, note that ${\sf lhead}$ and ${\sf ltail}$ are total functions for well-formed regular expressions to well-formed regular expressions.
%
%In the S-T-Concat case, note that Lemma~\ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.
%
%In the S-T-Replace case, it suffices to show that $\lreplace{r}{r_1}{r_2}$ is a regular expression assuming (inductively) that $r, r_1$ and $r_2$ are all regular expressions.
%This follows from the Closure proposition.
%\end{proof}

@@

%\begin{trconj}
%if $\Gamma \vdash e : \sigma$ then $e \mapsto^* v$ such that $v \ \val$.
%\end{trconj}
%%\begin{trthm}[Termination of small step semantics]
%%\end{trthm}
%%\begin{proof}
%We do not develop the full proof here, but note that the simply typed lambda calculus terminates.
%For the string fragment, observe that the S-T- rules do not add any non-trivial binding structure because substitutions
%$[e/x]e'$ may only occur in the special case where $e = \rstr{s}$, so that the length of the term
%never increases and the number of free variables strictly decreases. Therefore, the standard normalization
%argument proceeds without complication after fixing an evaluation order for the compatibility rules 
%(all our other proofs are agnostic to evaluation order).
%\end{proof}

\begin{trlem}[Canonical Forms]\label{cfgen}
If $\emptyset \vdash v : \sigma$ then:
\begin{enumerate}
\item If $\sigma = \stringin{r}$ then $u = \rstr{s}$.
\item If $\sigma = \sigma_1 \rightarrow \sigma_2$ then $v = \lambda x . e'$ and $s \in \lang{r}$.
\end{enumerate}

\end{trlem}
\begin{proof}
By inspection of valuation and typing rules.
\end{proof}

@@

\begin{lem}[Canonical Forms for the String Fragment of $\lambdas$]\label{thm:cfs}
  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
\end{lem}
\begin{proof}
This fact follows directly from Lemma~\ref{cfgen} and the preservation lemma.
\end{proof}

\begin{trlem}[Progress of small step semantics.]
If $\emptyset\vdash e : \sigma$ either $e = v$ or $e \mapsto e'$ for some $e'$.
\end{trlem}
\begin{proof}
The proof proceeds by induction on the derivation of $\emptyset\vdash e : \sigma$.
\begin{itemize}[label=$ $,itemsep=1ex]
\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of progress for
the simply typed lambda calculus.

\item \textbf{S-T-Stringin-I}. Suppose $\emptyset\vdash \rstr{s} : \stringin{s}$.
Then $\rstr{s} \ \val$ by SS-E-RStr.

\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{r}$ and $\emptyset e_1 : \stringin{r_1}$ and $\emptyset \vdash e_2 : \stringin{r_2}$.
By induction, $e_1 \mapsto e_1'$ or $e_1 = v_1$ and similarly
for $e_2$. 
If $e_1$ steps, then SS-E-Concat-Left applies and so
$\rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$. Similarly, if $e_2$ steps
then $e$ steps by SS-E-Concat-Right.

In the remaining case, $e_1 =v_1$ and $e_2 =v_2$.
But then it follows by Canonical Forms that $e_1 = \rstr{s_1}$
and $e_2 = \rstr{s_2}$.
Finally, by SS-E-Concat, $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.

\def\mline#1{\par\hspace*{-\leftmargin}\parbox{\textwidth}{\[#1\]}}
\item \textbf{S-T-Case}.
Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$ and $\emptyset\vdash e_1 : \stringin{r}$.
%\begin{align}
%&\vdash e_1 : \stringin{r} \\
%&\vdash e_2 : \sigma \\
%x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma
%\end{align}
%Actually this isn't even necessary...
%By (2), $e_2 \mapsto e_2'$ or $e_2 \val$. 
%In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
By induction and Lemma~\ref{thm:cfs} it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
In the former case, $e$ steps by S-E-Case-Left.
In the latter case, note that $s = \epsilon$ or $s = at$ for some string $t$.
If $s = \epsilon$ then $e$ steps by S-E-Case-$\epsilon$-Val,
and if $s = at$ the $e$ steps by S-E-Case-Concat.

\item \textbf{S-T-Replace}.
Suppose $e = \rreplace{r}{e_1}{e_2}$, $\emptyset\vdash e : \stringin{r'}$ and:
\begin{align}
&\emptyset\vdash e_1 : \stringin{r_1} \\
&\emptyset\vdash e_2 : \stringin{r_2} \\
&\lreplace{r}{r_1}{r_2} = r'
\end{align}

By induction on (1), $e_1 \mapsto e_1'$ or $e_1 = v_1$  for some $e_1'$.
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-Replace-Left.
Similarly, if $e_2$ steps then $e$ steps by SS-E-Replace-Right.
The only remaining case is where $e_1 = v_1$ and also $e_2 = v_2$.
By Canonical Forms, $e_1 = \strin{s_1}$ and $e_2 = \strin{s_2}$.
Therefore, $e \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ by SS-E-Replace.

\item \textbf{S-T-SafeCoerce}.
Suppose that $\emptyset\vdash \rcoerce{r}{e_1} : \stringin{r}$.
and
%\begin{align}
$\emptyset\vdash e_1 : \stringin{r'}$ for $\lang{r'} \subseteq \lang{r}$.% \\
%&\lang{r'} \subseteq \lang{r}
%\end{align}
By induction, $e_1 =v_1$ or $e_1 \mapsto e_1'$ for some $e_1'$.
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-SafeCoerce-Step.
Otherwise, $e_1 =v$ and by Canonical Forms $e_1 = \rstr{s}$.
In this case, $e = \rcoerce{r}{\rstr{s}} \mapsto \rstr{s}$ by SS-E-SafeCoerce.

\item \textbf{S-T-SafeCheck}
Suppose that $\emptyset\vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \stringin{r}$ and:
\begin{align}
&\vdash e_0 : \stringin{r_0}  \\
x:\stringin{r} &\vdash e_1 : \sigma \\
&\vdash e_2 : \sigma \label{align6}
\end{align}

By induction,  $e_0 \mapsto e_0'$ or $e_0 = v$.
In the former case $e$ steps by SS-E-Check-StepRight.
Otherwise, $e_0 = \rstr{s}$ by Canonical Forms.
By Lemma~\ref{thm:regexprops} part 2, either $s \in \lang{r_0}$ or $s \not\in\lang{r_0}$.
In the former case $e$ takes a step by SS-E-Check-Ok.
In the latter case $e$ takes a step by SS-E-Check-NotOk.

@@

\begin{trlem}[Substitution] \label{lem:subst}
If $\Psi\ x : \sigma \vdash e : \sigma'$ and $\psi \vdash e' : \sigma$, then
$\psi \vdash [e' / x] e : \sigma'$.
\end{trlem}

\begin{trlem}[Preservation for Small Step Semantics]
If $\emptyset \vdash e : \sigma$ and $e \mapsto e'$ then $\emptyset \vdash e' : \sigma$.
\end{trlem}
\begin{proof}

By induction on the derivation of $e \mapsto e'$ and $\emptyset \vdash e : \sigma$.

@@

\item \textbf{SS-E-Concat-Left}.
Suppose $e = \rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$ and $e_1 \mapsto e_1'$.
The only rule that applies is S-T-Concat, so $\emptyset\vdash e_1 : \stringin{r_1}$ and $\emptyset\vdash e_2 : \stringin{r_2}$.
By induction, $\emptyset\vdash e_1' : \stringin{r_1}$.
Therefore, by S-T-Concat, $\emptyset\vdash \rconcat{e_1'}{e_2} : \stringin{r_1r_2}$.

\item {\textbf{SS-E-Concat-Right}}. 
Suppose $e = \rconcat{e_1}{e_2} \mapsto \rconcat{e_1}{e_2'}$ and $e_2 \mapsto e_2'$.
The only rule that applies is S-T-Concat, so $\emptyset\vdash e_1 : \stringin{r_1}$ and $\emptyset\vdash e_2 : \stringin{r_2}$.
By induction, $\emptyset\vdash e_2' : \stringin{r_2}$.
Therefore, by S-T-Concat, $\emptyset\vdash \rconcat{e_1}{e_2'} : \stringin{r_1r_2}$.

\item \textbf{SS-E-Concat}. 
Suppose $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
The only applicable rule is S-T-Concat, so 
$\emptyset\vdash\rstr{s_1} : \stringin{r_1}$ 
and
$\emptyset\vdash\rstr{s_2} : \stringin{r_2}$. 
By Lemma~\ref{lem:cfs}, $s_1 \in \lang{r_1}$ and $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
that $s_1s_2 \in \lang{r_1r_2}$. Therefore, $\emptyset\vdash \rstr{s_1s_2} : \stringin{r_1r_2}$ by S-T-Rstr.

\item \textbf{S-E-Case-Left}.
Suppose 
$e \mapsto \strcase{e_1'}{e_2}{x,y.e_3}$
and $\emptyset \vdash e : \sigma$ and $e_1 \rightarrow e_1'$.
The only rule that applies is S-T-Case, so:
\begin{align}
\emptyset&\vdash e_1 : \stringin{r} \label{align7} \\
\emptyset&\vdash e_2 : \sigma \label{align8} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align9}
\end{align}
By (\ref{align7}) and the assumption that $e_1 \mapsto e_1'$, it follows by induction that $\emptyset\vdash e_1' : \stringin{r}$.
This fact together with (\ref{align8}) and (\ref{align9}) implies by S-T-Case that
$\emptyset\vdash\strcase{e_1'}{e_2}{x,y.e_3} : \sigma$.
